---
title: "Practical 2: ggplot2"
output: html_notebook
---

#to install gcookbook and load it
```{r}
library(gcookbook)
library(dplyr)
library(ggplot2)
load("./doi_10.15146_mdpr-pm59__v1/marena_tree_17Jul2019/marena1cns.tree1.rdata")
```

```{r}
ggplot(data = marena1cns.tree1, aes(x = ba, y= agb)) +
  geom_point() 
 
```


```{r}
plot1 <- ggplot(marena1cns.tree1, aes(x = ba, y= agb)) +
  geom_point() 

 plot1 

```

```{r}
ggplot(marena1cns.tree1, aes(x = ba, y= agb)) 
```


```{r}
filtered <- filter(marena1cns.tree1,plot == "p17", ba > 0, sp %in% c("brosco", "cha1te", "poular", "oxanlo"))

filtered
 
```


###Example Code 2
```{r}

 ggplot(filtered, aes(x = ba, y= agb)) +
  geom_point() 

```



###Example code 3
```{r}
ggplot(heightweight, aes(x = ageYear, y = heightIn, shape = sex, colour = sex)) +
  geom_point()
```


###Task 1
###A.	Using Example Code 3 as an example, modify example code 2  so the data from each species are represented in a different colour
###B.	Now modify your scatter plot so that you can make a line plot instead (you can find example code in the r graphics cookbook)
###C.	Now modify your line plot so you only plot the lower left hand corner. To do this add the example code 4 below to your existing code and modify the limits if necessary.


###Answer 1A



```{r}
ggplot(filtered, aes(x = ba, y= agb, colour = sp)) +
  geom_point() 

```


###Answer 1B
```{r}
 ggplot(filtered, aes(x = ba, y= agb, colour = sp)) +
  geom_line() 
```


###Example code 4
```{r}
  #         +
  #coord_cartesian(ylim = c(0, 1), xlim = c(0,1))

```



```{r}
 ggplot(filtered, aes(x = ba, y= agb, colour = sp)) +
  geom_line()  +
  coord_cartesian(ylim = c(0, 0.1), xlim = c(0,0.03))
```







```{r}
ce_mod <- cabbage_exp %>%
  filter(Cultivar == "c39")
ce_mod
```
 
 

###Example Code 5 
```{r}

ggplot(ce_mod, aes(x = Date, y = Weight)) +
  geom_point(size = 4) +
  geom_errorbar(aes(ymin = Weight - se, ymax = Weight + se), width = .2)
 
```
 
###Task 2
###A.	Using what you learned in prac 1, summarise the filtered datset to produce a summary dataset with new variables for mean of agb and standard deviation of agb. This is a revision exercise. If you want to skip this or if you get stuck, there is code at the end of this practical you can use
###B.	Using example code 4 and the explanation above, make an error bar chart for your new summary data


#The need to summarise their data (I have supplied this code at the end of the pra instructions)
```{r}
filtered.sum <- filtered %>%
group_by(sp) %>%
summarise(ave.mass = mean(agb), sd = sd(agb))

```

###Answer 2B
```{r}
ggplot(filtered.sum, aes(x = sp, y = ave.mass)) +
  geom_point(size = 4) +
  geom_errorbar(aes(ymin = ave.mass - sd, ymax = ave.mass + sd), width = .2)

```



###Task 3
###A.	Using the example here, add some more descriptive axis labels to the x and y axes of the plot from task 2B 
###B.	Find instructions in the r graphics cookbook and add a title to your plot
###C.	Read this section of the cookbook and change the angles of the text on the x-axis
###D.	Find instructions in the r graphics cookbook and reverse the order of the species on the x-axis


###Answer 3A
```{r}
ggplot(filtered.sum, aes(x = sp, y = ave.mass)) +
  geom_point(size = 4) +
  geom_errorbar(aes(ymin = ave.mass - sd, ymax = ave.mass + sd), width = .2) +
  xlab("Species") +
  ylab("Average Above Ground Biomass")
  
```

###Answer 3B
###https://r-graphics.org/recipe-appearance-title
```{r}
ggplot(filtered.sum, aes(x = sp, y = ave.mass)) +
  geom_point(size = 4) +
  geom_errorbar(aes(ymin = ave.mass - sd, ymax = ave.mass + sd), width = .2)+
  ggtitle("Biomass of selected tree species from Panama")
```
###Answer 3C
```{r}
ggplot(filtered.sum, aes(x = sp, y = ave.mass)) +
  geom_point(size = 4) +
  geom_errorbar(aes(ymin = ave.mass - sd, ymax = ave.mass + sd), width = .2) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1))
```


###Answer 3D
###https://r-graphics.org/recipe-axis-order
```{r}
ggplot(filtered.sum, aes(x = sp, y = ave.mass)) +
  geom_point(size = 4) +
  geom_errorbar(aes(ymin = ave.mass - sd, ymax = ave.mass + sd), width = .2) +
  scale_x_discrete(limits = c("poular", "oxanlo", "cha1te", "brosco"))
```






